Placing the spotted T Tauri star LkCa 4 on an HR diagram

### 2016-08-03 document/ms.tex
Use Grammarly to find grammar and spelling mistakes.  Awkward with latex though.

In this work we constrain the starspot properties of \name with 4 complementary techniques.  Section \ref{sec:obs} quantifies the optical variability of \name over the last 31 years based on all available photometric monitoring; a collection of spectral observations at various phases of variability is introduced.  Section \ref{sec:methods} and Appendix \ref{methods-details} describe a spectral inference technique aimed at generalizing line depth ratio analysis, which is applied to a high resolution panchromatic near-IR spectrum of \name in Section \ref{sec:two_tempIGRINS} .  Sections \ref{sec:SED}, \ref{sec:rotSpot}, and \ref{sec:RotTiO} show results from SED fitting, polychromatic photometric monitoring, and optical TiO line-depth ratio fitting.  All lines of evidence are ultimately combined to build a consistent picture of the spectral and temporal evolution of \name, and what can be understood about pre-MS stellar evolution from this exemplar.

@@

\subsection{ESPaDOnS Spectroscopy}
We used ESPaDOnS on CFHT to obtain twelve high resolution optical spectra of \name from 8-21 Jan.~2014 as part of the MATYSSE Large Program.  These spectra cover 3900$-$10000 \AA\ at $R\sim68,000$ and were obtained in spectropolarimetry mode.  The Zeeman Doppler Imaging obtained from these observations were analyzed by \citet{donati14}.  In this paper we concentrate on the intensity spectrum.  
%\todo{Flux calibration?}

@@

An archive of 1216 $B$, $V$ and/or $R$ photometric monitoring visits was assembled from published and unpublished data sources spanning 31 years in 22 observing seasons.  Published data targeting \name is composed of 29 $BVR$ visits from 1985$-$1986 \citep{vrba93}, 26 $BVR$ epochs from 1990$-$1991 \citep{bouvier93}, 284 visits in $V$-band (278 with $B$ and 268 with $R$) from 1992 August - 2004 October \citep{grankin08}, and 10 $BVR$ visits from 2013 \citep{donati14}.  ASAS3 \citep{pojmanski04} acquired 63 $V-$band measurements from 2002$-$2004.  Integral-OMC obtained 138 $V-$band measurements from 2006$-$2008 \citep{garzon12}.  The AAVSO archive \citep{kafka16} includes 385 $V$, 23 $B$ and 10 $R$ measurements from 2013$-$2016.  Unpublished data from the ASAS-SN survey \citep{shappee14}, were obtained from 2012 January -- 2016 March in 186 visits.  Finally, recent photometry from the ongoing Crimean Astrophysical Observatory (CrAO) ROTOR project \citep{grankin08} is presented here for the first time.  The CrAO data includes 43 $BVR$ visits from August 2015 - March 2016. Figure \ref{fig:PhotTime} displays all the $V-$band photometry over the interval 1985-2016.  HJD, BJD, and JD are simply refered to as JD, which is accurate to the precision of available data.

@@

As noted previously, line depth ratio analysis has traditionally been limited to isolated portions of spectrum that possess spectral lines attributable only to starspots and spectral lines attributable only to ambient photosphere \citep[\emph{e.g.}][]{neff95, oneal01}.  The apparent veiling of the lines and their respective temperature dependences can be combined to solve for the starspot and ambient photosphere temperatures and relative areal coverages.  The line depth ratio method suffers from the need to identify portions of the spectrum that possess such strong lines, and also suffers from the need to assemble large atlases of observed spectral templates to which the spotted star spectrum can be compared.  In this section we introduce a generalization of line depth ratio analysis which employs pixel-by-pixel modeling of spectral orders.  By leveraging all the channels present in a spectrum, this strategy has the power to constrain starspot properties at relatively low fill-factors, and identify weak lines that originate from the starspot.

@@

Note that $f_{\Omega}$ represents an instantaneous, observational fill factor seen on one projected hemisphere, not $f_{\rm spot}$, the ``ratio of the spotted surface to the total surface areas'' \citep{somers15}.  In the limit of homogenously distributed spots, $f_{\Omega}$ tends to $f_{\rm spot}$, but in general $f_{\rm spot}$ is not a direct observable since some circumpolar regions of inclined stars will always face away from Earth; $f_{\Omega}$ can vary cyclicly though rotational modulation, whereas $f_{\rm spot}$ changes secularly through starspot evolution that is not yet understood.

@@

We find a larger dispersion in the point estimates for the stellar parameters derived from the IGRINS data than those derived in optical.  The most conspicuous trend is in the derived effective temperature as a function of wavelength shown in Figure \ref{fig:SingleTeffvsOrder}.  The effective temperature peaks at a values of $\sim4200$K in the short wavelength end of $H-$band and saturates at $<3500$ K at the long wavelength end of $H-$band.  The $K-$band shows even lower derived effective temperatures of $\sim3300$ K, a $Delta \teff \sim 700$ K spread from optical.  No single temperature can describe all the spectral lines present in the high resolution optical and IR spectra.  This discrepancy is circumstantial evidence for the detection of spectral features attributable to starspots.

@@

The heightened sensitivity to starspot spectral lines as a function of wavelength can be understood in the following way.  Starspots are cooler than their surrounding photosphere and will therefore have a longer wavelength of peak emission, according to Wein's displacement law.  The average ratio of flux density between starspot and bulk photosphere will increase with wavelength until asymptoting to a fixed value in the Rayleigh-Jeans tail \citep{wolk96}.

In summary, the visible portion of the spectrum will be dominated by the warm patches, but the contrast with the starspots will be higher.  The infrared spectrum will have a lower overall spot contast, but will have more net starspot flux than optical.  This observation has motivated precision radial velocity (RV) planet searches in the infrared, where the modulation of RV signals is lessened \citep{prato08}.

@@

In the previous sections we established that the high resolution optical and near-IR spectra of LkCa 4 may be explained by a cool photosphere with a hot spot.  These results have testable predictions for the spectral energy distribution (SED) and rotational modulation of features of the SED.

@@

Extinction paragraph, needs overhaul...  Relative to main sequence colors, the $E(B-V)=0.24$ would yield $A_V=0.72$ mag....  
The $J-K=0.93$ color is equal to the expected color to within 0.05 => 0.3 mag (for T=3000, 4200)... The $B-V$ is always dominated by the hot component and therefore does not depend on the filling factor of the hot component.  The $V-R$ varies more with $V$, indicating a larger contribution from the cool component.  The difference between observed and predicted $B-V$ color, $E(B-V)$, may correspond to the extinction.  After accounting for extinction synthetic model underpredicts both the $V-R$...

@@

The optical emission is dominated by the hotter component while both components contribute to the red and infrared emission (see Figure \ref{fig:lores}).  The $0.7$ mag range in $V$-band brightness requires at a factor of 1.9 change in the visible surface area of hot component in the limit that the cool component does not contribute any $V$-band emission.  In the 2014-2015 ASAS-SN photometry, the minimum and maximum brightness correspond to cool spot coverage of 83\% and 68\%, respectively.  LkCa 4 was brighter in past epochs, which corresponds to either a larger covering fraction for the hot component (as assumed here) or components with hotter temperatures.

@@

\begin{figure*}
	\centering
	\includegraphics[trim=2.6cm 13.0cm 2.8cm 3.5cm, clip=true, width=0.65\textwidth]{figures/lkca4_hrdiag_forpaper}
	\caption{Locations for LkCa 4 on an HR diagram, compared with models of pre-main sequence evolution calculated by \citet{baraffe15} with isochrones (black lines) and evolution models of a single mass (dashed blue lines) as marked.  The measured effective temperature and luminosity from this paper, based on the two-component fit and a median $V$-band magnitude, corresponds to the black asterisk.  The yellow shaded region corresponds to the range of apparent effective temperatures that are would be measured as the hot component rotates into and out of our view.  The blue circle corresponds to the measurement at blue-optical wavelengths by \citet{donati14}, the purple square corresponds to the measurement from low-resolution optical spectra, biased to TiO band strengths, by \citet[][biased to]{herczeg14}, while the red diamong corresponds to what we would measure from the K-band spectrum and 2MASS $J$-band magnitude.}
	\label{fig:hrdiag}
\end{figure*}

@@

\emph{Number of photospheric components-}  Our full spectrum fitting method assumed \emph{only} two temperature components present in the photosphere.  In reality, starspots are probably described by a range of effective temperatures like umbra, penumbra, and plages on the sun's surface.  A model possessing three unique temperatures and two unique fill factors might better approximate the minor constituents of the emergent spectrum than our two temperature model can.  Computational infeasibility limited our ability to rerun new multi-temperature models and compute model comparison diagnostics like AIC, BIC, or cross validation \citep{ivezic14}.  Such investigations could be considered in future works, but we suspect systematic errors in the models would ultimately limit the interpretability of models with more than two temperature components.  High-precision polychromatic planetary transit photometry could offer an avenue to measure the morphologies and temperatures of starspots for the subset of transiting exoplanet host stars.
% The fact that we get consistent estimates for Teff tells us something.
% The fits are pretty good, also tells us something.
% Coarsely sampled models + spectral emulator acts as a smoother in Teff, as does the GP global covariance kernel.
%(cite Charlotte Norris CS19 poster?)

\emph{Suitability of pre-computed \PHOENIX\ model spectra, model ranges, $\A \neq 0$, and magnetic fields-} We have assumed that the \PHOENIX\ synthetic spectra are a good representation to our observed IGRINS spectrum of \name, a young star likely to possess a large magnetic field.  Cool sunspots are associated with heightened magnetic fields on the sun.  Magnetic fields will have two effects on the emergent spectrum- line broadening attributable to the Zeeman effect, and pressure broadening attributable to seeing deeper into deeper layers in the photosphere compared to the surrounding hotter photosphere.  The former effect will differentially impact lines with high Lande-g factors (cite XX, fix the g formatting).  The latter effect could masquerade as an apparent shift in surface gravity, as seen in mixture models of spotted stars with empirical templates composed of dwarfs and giants (cite XX Saar and collaborators).  Our tacit assumption has been that these effects are 1) relatively small compared to the gross appearance and disappearance of spectral lines at disparate temperatures and 2) there exist many orders and lines that are not affected at all.  We see tentative evidence for physics outside of the pre-computed model grids in the poor fits to the observed spectra in $K-$band.  Strong metal lines demonstrate large departures in their best fit $\Z$ due to Zeeman broadening.  No pre-computed model grids including a range of surface magnetic fields are available at the wavelength range and spectral sampling needed to approach these questions within this spectral inference framework.
% Note that these two effects would differently affect the spectra

We also assumed that the spectral pattern of metallicity can be captured by a single coarse metric, $\Z$.  In reality, the low temperature starspot spectrum of \name could exhibit $\A \neq 0$, resulting in a greater amplitude of residual spectrum correlations, and/or biased estimates for other parameters.  Our Gaussian process inference framework yields some resilience to the former problem.  For the latter, nonzero $\A$ would only impact a subset of all lines, and only to at a magnitude of typical departures of $\A$ in the solar neighborhood.  The large number of IGRINS orders and spectral lines offers us some resilience to isolated orders badly affected by spectral line outliers.

>>>>

In this work, we constrain the starspot properties of \name with 4 complementary techniques.  Section \ref{sec:obs} quantifies the optical variability of \name over the last 31 years based on all available photometric monitoring; a collection of spectral observations at various phases of variability is introduced.  Section \ref{sec:methods} and Appendix \ref{methods-details} describe a spectral inference technique aimed at generalizing line depth ratio analysis, which is applied to a high resolution panchromatic near-IR spectrum of \name in Section \ref{sec:two_tempIGRINS} .  Sections \ref{sec:SED}, \ref{sec:rotSpot}, and \ref{sec:RotTiO} show results from SED fitting, polychromatic photometric monitoring, and optical TiO line-depth ratio fitting.  All lines of evidence are ultimately combined to build a consistent picture of the spectral and temporal evolution of \name, and what can be understood about pre-MS stellar evolution from this exemplar.

@@

\subsection{ESPaDOnS Spectroscopy}
We used ESPaDOnS on CFHT to obtain twelve high resolution optical spectra of \name from 8-21 Jan.~2014 as part of the MATYSSE Large Program.  These spectra cover 3900$-$10000 \AA\ at $R\sim68,000$ and were obtained in spectropolarimetry mode.  The Zeeman Doppler Imaging obtained from these observations were analyzed by \citet{donati14}.  In this paper, we concentrate on the intensity spectrum.  
%\todo{Flux calibration?}

@@

An archive of 1216 $B$, $V$ and/or $R$ photometric monitoring visits was assembled from published and unpublished data sources spanning 31 years in 22 observing seasons.  Published data targeting \name is composed of 29 $BVR$ visits from 1985$-$1986 \citep{vrba93}, 26 $BVR$ epochs from 1990$-$1991 \citep{bouvier93}, 284 visits in $V$-band (278 with $B$ and 268 with $R$) from 1992 August - 2004 October \citep{grankin08}, and 10 $BVR$ visits from 2013 \citep{donati14}.  ASAS3 \citep{pojmanski04} acquired 63 $V-$band measurements from 2002$-$2004.  Integral-OMC obtained 138 $V-$band measurements from 2006$-$2008 \citep{garzon12}.  The AAVSO archive \citep{kafka16} includes 385 $V$, 23 $B$ and 10 $R$ measurements from 2013$-$2016.  Unpublished data from the ASAS-SN survey \citep{shappee14} were obtained from 2012 January -- 2016 March in 186 visits.  Finally, recent photometry from the ongoing Crimean Astrophysical Observatory (CrAO) ROTOR project \citep{grankin08} is presented here for the first time.  The CrAO data includes 43 $BVR$ visits from August 2015 - March 2016. Figure \ref{fig:PhotTime} displays all the $V-$band photometry over the interval 1985-2016.  HJD, BJD, and JD are simply referred to as JD, which is accurate to the precision of available data.

@@

As noted previously, line depth ratio analysis has traditionally been limited to isolated portions of spectrum that possess spectral lines attributable only to starspots and spectral lines attributable only to ambient photosphere \citep[\emph{e.g.}][]{neff95, oneal01}.  The apparent veiling of the lines and their respective temperature dependences can be combined to solve for the starspot and ambient photosphere temperatures and relative areal coverages.  The line depth ratio method suffers from the need to identify portions of the spectrum that possess such strong lines, and also suffers from the need to assemble large atlases of observed spectral templates to which the spotted star spectrum can be compared.  In this section, we introduce a generalization of line depth ratio analysis which employs pixel-by-pixel modeling of spectral orders.  By leveraging all the channels present in a spectrum, this strategy has the power to constrain starspot properties at relatively low fill-factors and identify weak lines that originate from the starspot.

@@

Note that $f_{\Omega}$ represents an instantaneous, observational fill factor seen on one projected hemisphere, not $f_{\rm spot}$, the ``ratio of the spotted surface to the total surface areas'' \citep{somers15}.  In the limit of homogenously distributed spots, $f_{\Omega}$ tends to $f_{\rm spot}$, but in general $f_{\rm spot}$ is not a direct observable since some circumpolar regions of inclined stars will always face away from Earth; $f_{\Omega}$ can vary cyclically through rotational modulation, whereas $f_{\rm spot}$ changes secularly through starspot evolution that is not yet understood.

@@

We find a larger dispersion in the point estimates for the stellar parameters derived from the IGRINS data than those derived in optical.  The most conspicuous trend is in the derived effective temperature as a function of wavelength shown in Figure \ref{fig:SingleTeffvsOrder}.  The effective temperature peaks at values of $\sim4200$K in the short wavelength end of $H-$band and saturates at $<3500$ K at the long wavelength end of $H-$band.  The $K-$band shows even lower derived effective temperatures of $\sim3300$ K, a $Delta \teff \sim 700$ K spread from optical.  No single temperature can describe all the spectral lines present in the high resolution optical and IR spectra.  This discrepancy is circumstantial evidence for the detection of spectral features attributable to starspots.

@@

The heightened sensitivity to starspot spectral lines as a function of wavelength can be understood in the following way.  Starspots are cooler than their surrounding photosphere and will, therefore, have a longer wavelength of peak emission, according to Wein's displacement law.  The average ratio of flux density between starspot and bulk photosphere will increase with wavelength until asymptoting to a fixed value in the Rayleigh-Jeans tail \citep{wolk96}.

In summary, the visible portion of the spectrum will be dominated by the warm patches, but the contrast with the starspots will be higher.  The infrared spectrum will have a lower overall spot contrast, but will have more net starspot flux than optical.  This observation has motivated precision radial velocity (RV) planet searches in the infrared, where the modulation of RV signals is lessened \citep{prato08}.

@@

In the previous sections, we established that the high resolution optical and near-IR spectra of LkCa 4 may be explained by a cool photosphere with a hot spot.  These results have testable predictions for the spectral energy distribution (SED) and rotational modulation of features of the SED.

@@

Extinction paragraph, needs overhaul...  Relative to main sequence colors, the $E(B-V)=0.24$ would yield $A_V=0.72$ mag....  
The $J-K=0.93$ color is equal to the expected color to within 0.05 => 0.3 mag (for T=3000, 4200)... The $B-V$ is always dominated by the hot component and therefore does not depend on the filling factor of the hot component.  The $V-R$ varies more with $V$, indicating a larger contribution of the cool component.  The difference between observed and predicted $B-V$ color, $E(B-V)$, may correspond to the extinction.  After accounting for extinction synthetic model underpredicts both the $V-R$...

@@

The optical emission is dominated by the hotter component while both components contribute to the red and infrared emission (see Figure \ref{fig:lores}).  The $0.7$ mag range in $V$-band brightness requires a factor of 1.9 change in the visible surface area of the hot component in the limit that the cool component does not contribute any $V$-band emission.  In the 2014-2015 ASAS-SN photometry, the minimum and maximum brightness correspond to cool spot coverage of 83\% and 68\%, respectively.  LkCa 4 was brighter in past epochs, which corresponds to either a larger covering fraction for the hot component (as assumed here) or components with hotter temperatures.

@@

\begin{figure*}
 \centering
 \includegraphics[trim=2.6cm 13.0cm 2.8cm 3.5cm, clip=true, width=0.65\textwidth]{figures/lkca4_hrdiag_forpaper}
 \caption{Locations for LkCa 4 on an HR diagram, compared with models of pre-main sequence evolution calculated by \citet{baraffe15} with isochrones (black lines) and evolution models of a single mass (dashed blue lines) as marked.  The measured effective temperature and luminosity from this paper, based on the two-component fit and a median $V$-band magnitude, corresponds to the black asterisk.  The yellow shaded region corresponds to the range of apparent effective temperatures that are would be measured as the hot component rotates into and out of our view.  The blue circle corresponds to the measurement at blue-optical wavelengths by \citet{donati14}, the purple square corresponds to the measurement from low-resolution optical spectra, biased to TiO band strengths, by \citet[][biased to]{herczeg14}, while the red diamond corresponds to what we would measure from the K-band spectrum and 2MASS $J$-band magnitude.}
 \label{fig:hrdiag}
\end{figure*}

@@

\emph{Number of photospheric components-}  Our full spectrum fitting method assumed \emph{only} two temperature components present in the photosphere.  In reality, starspots are probably described by a range of effective temperatures like umbra, penumbra, and plages on the sun's surface.  A model possessing three unique temperatures and two unique fill factors might better approximate the minor constituents of the emergent spectrum than our two temperature model can.  Computational infeasibility limited our ability to rerun new multi-temperature models and compute model comparison diagnostics like AIC, BIC, or cross-validation \citep{ivezic14}.  Such investigations could be considered in future works, but we suspect systematic errors in the models would ultimately limit the interpretability of models with more than two temperature components.  High-precision polychromatic planetary transit photometry could offer an avenue to measure the morphologies and temperatures of starspots for the subset of transiting exoplanet host stars.
% The fact that we get consistent estimates for Teff tells us something.
% The fits are pretty good, also tells us something.
% Coarsely sampled models + spectral emulator acts as a smoother in Teff, as does the GP global covariance kernel.
%(cite Charlotte Norris CS19 poster?)

\emph{Suitability of pre-computed \PHOENIX\ model spectra, model ranges, $\A \neq 0$, and magnetic fields-} We have assumed that the \PHOENIX\ synthetic spectra are a good representation of our observed IGRINS spectrum of \name, a young star likely to possess a large magnetic field.  Cool sunspots are associated with heightened magnetic fields on the sun.  Magnetic fields will have two effects on the emergent spectrum- line broadening attributable to the Zeeman effect, and pressure broadening attributable to seeing deeper into deeper layers in the photosphere compared to the surrounding hotter photosphere.  The former effect will differentially impact lines with high Lande-g factors (cite XX, fix the g formatting).  The latter effect could masquerade as an apparent shift in surface gravity, as seen in mixture models of spotted stars with empirical templates composed of dwarfs and giants (cite XX Saar and collaborators).  Our tacit assumption has been that these effects are 1) relatively small compared to the gross appearance and disappearance of spectral lines at disparate temperatures and 2) there exist many orders and lines that are not affected at all.  We see tentative evidence for physics outside of the pre-computed model grids in the poor fits to the observed spectra in $K-$band.  Strong metal lines demonstrate large departures in their best fit $\Z$ due to Zeeman broadening.  No pre-computed model grids including a range of surface magnetic fields are available at the wavelength range and spectral sampling needed to approach these questions within this spectral inference framework.
% Note that these two effects would differently affect the spectra

We also assumed that the spectral pattern of metallicity can be captured by a single coarse metric, $\Z$.  In reality, the low-temperature starspot spectrum of \name could exhibit $\A \neq 0$, resulting in a greater amplitude of residual spectrum correlations, and/or biased estimates for other parameters.  Our Gaussian process inference framework yields some resilience to the former problem.  For the latter, nonzero $\A$ would only impact a subset of all lines, and only to at a magnitude of typical departures of $\A$ in the solar neighborhood.  The large spectral grasp of IGRINS offers us some resilience to isolated orders badly affected by spectral line outliers.
